---
title: "Multivariate Distributions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multivariate Distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
has_mvtnorm <- requireNamespace("mvtnorm", quietly = TRUE)
```

```{r setup}
library(distionary)
```

A distribution can describe several variables at once: the flows of two rivers on the same day, say, or a storm's duration together with its depth. This vignette shows how to build such distributions, how to evaluate them, and how to condition them --- including slicing one along a line, such as fixing the total of two flows.

Multivariate distributions are new in `distionary`, and still experimental. The names and behaviour described here may change.

## Building one

The multivariate Normal comes in two forms: `dst_mv_norm()` for any number of variables, specified by a mean vector and covariance matrix, and `dst_bi_norm()`, a shortcut for two, specified by standard deviations and a correlation.

```{r}
d <- dst_bi_norm(mean = c(0, 1), sd = c(1, 2), cor = 0.6)
d
```

Its variables have names, and `dimension()` counts them.

```{r}
variables(d)
dimension(d)
```

Name them yourself through the names of the mean:

```{r}
flows <- dst_mv_norm(
  mean = c(r1 = 50, r2 = 80),
  cov = matrix(c(100, 60, 60, 225), 2)
)
variables(flows)
```

The other built-in family is the empirical distribution, which places probability on each observed point --- each row of a data set, taken as a whole.

```{r}
obs <- data.frame(
  flow = c(10, 12, 15, 11, 12),
  depth = c(1.0, 1.5, 2.0, 1.1, 1.5)
)
e <- dst_bi_empirical(flow, depth, data = obs)
e
```

`dst_mv_empirical()` does the same for any number of variables: `dst_mv_empirical(flow, depth, data = obs)`, or, to take many columns at once, `dst_mv_empirical(obs[cols])`.

There is also the multivariate Student t, `dst_mv_t()` (and `dst_bi_t()`), whose heavier tails make its variables extreme together more often than the Normal's. Everything below that works for the Normal --- including slicing --- works for it too.

## Evaluating one

A univariate distribution is evaluated at a vector of values. A multivariate one needs a value for each variable, so there are two families of evaluators, much as `purrr` has `map2()` and `pmap()`:

- `eval_bi_*(d, x, y)` for two variables, one vector each;
- `eval_mv_*(d, l)` for any number, a list of vectors.

```{r, eval = has_mvtnorm}
eval_bi_cdf(d, x = 0, y = 1)
eval_mv_cdf(d, list(0, 1))
eval_bi_density(d, x = 0:2, y = 1)
```

The vectors are recycled to a common length, and element `i` of the result is the representation at the `i`th point. A named list is matched to the variables by name, so its order does not matter; a data frame works just as well. That includes the draws from `realise()`, which come back as a data frame:

```{r}
set.seed(1)
draws <- realise(flows, n = 3)
draws
eval_mv_density(flows, draws)
```

The functions for one variable refuse a multivariate distribution, and say what to use instead:

```{r, error = TRUE}
eval_cdf(d, 0)
```

### Survival, and other corners

The survival function is the probability that *every* variable exceeds its value, \(P(X > x, Y > y)\). That is not one minus the CDF, which is the probability that *at least one* does; the two agree only for a single variable.

```{r, eval = has_mvtnorm}
eval_bi_survival(d, x = 0, y = 1)
1 - eval_bi_cdf(d, x = 0, y = 1)
```

For the probability of any other event --- one variable below its value and the other above, say --- use `prob()`, described next.

## Probabilities of events

`prob()` finds the probability of an event, written as you would write it for `dplyr::filter()`: compare variables by name, and list the conditions that must all hold, separated by commas. Within a condition, `&` (and), `|` (or), and `!` (not) combine comparisons.

```{r, eval = has_mvtnorm}
prob(d, x <= 0, y > 1)
prob(d, x > 2 | y > 5)
```

A variable left out of the event is free, so there is no need to take a marginal first:

```{r}
prob(d, x < 2)
```

Values can be vectors, giving one probability for each:

```{r}
prob(d, y > c(1, 3, 5))
```

A comparison can involve several variables, as long as the distribution can say how their combination is distributed. The multivariate Normal and t can, for any sum of multiples of the variables:

```{r}
prob(d, x + y > 3)
prob(d, x < y)
```

For a distribution on finitely many points, anything goes: the event is checked at each point, much as `filter()` checks each row, and the probabilities of the points passing it are added up.

```{r}
prob(e, flow * depth > 15)
```

Strict and non-strict inequalities differ only where there are atoms, and are handled exactly:

```{r}
prob(e, flow < 12)
prob(e, flow <= 12)
```

`given` makes the probability conditional. Comparing a variable with `==` conditions on its value, which works even for a continuous variable, where that value has probability zero; any other condition is divided out as an event:

```{r}
prob(d, y > 3, given = x == 1)
prob(d, y > 3, given = x > 1)
```

A univariate distribution's variable is called `x`, so `prob()` works there too: `prob(dst_norm(0, 1), x > 2)`.

## Marginals and conditionals

`marginal()` gives the distribution of some of the variables, ignoring the rest. With one variable, it is an ordinary univariate distribution, to be evaluated with the usual functions.

```{r}
r2 <- marginal(flows, "r2")
r2
eval_quantile(r2, at = 0.99)
```

To evaluate a representation *conditional* on some of the variables, say which ones are known with `known`. The arguments stay in the same order either way: every variable still gets a value, and `known` says which of those values are known rather than evaluated at. The variables in `known` are the ones on the right of the bar.

```{r, eval = has_mvtnorm}
# P(Y <= y | X = 2), for y = 0, 1, 2
eval_bi_cdf(d, x = 2, y = 0:2, known = "x")
```

To have the conditional distribution itself, as a distribution to evaluate, draw from or condition further, use `conditional()` from the [distplyr](https://distplyr.probaverse.com) package, which makes new distributions out of existing ones.

## Supports of several variables

Every distribution says where it places its probability, through its support (see `vignette("support")`). The simplest multivariate support is a *product*: every value of one variable paired with every value of each of the others, as `expand.grid()` pairs them.

```{r}
support_product(
  storms = discrete(natural0()),
  depth = continuous(c(0, Inf))
)
```

An empirical distribution's support is not a product, though: it is the set of points observed, not every pairing of an observed flow with an observed depth. `discrete()` takes such a set as a data frame.

```{r}
support(e)
```

For anything else, describe the support as the image of a simpler one under a map. The triangle \(0 < y < x\) is where \((x, z) \mapsto (x, xz)\) sends the strip \(x > 0\), \(0 < z < 1\):

```{r}
strip <- support_product(
  x = continuous(c(0, Inf)),
  z = continuous(c(0, 1))
)
support_map(
  strip,
  fun = function(x, z) list(x, x * z),
  margins = list(
    x = continuous(c(0, Inf)),
    y = continuous(c(0, Inf))
  )
)
```

A general map has to be told what values each variable reaches. A linear one does not, and `support_affine()` works them out. Here is the line \(x + y = 100\), as the image of the real line under \(t \mapsto (50 + t, 50 - t)\):

```{r}
line <- support_affine(
  continuous(),
  shift = c(x = 50, y = 50),
  matrix = rbind(1, -1)
)
line
```

A line has two variables but spans only one dimension. A distribution on it puts no probability on any single point, but has no density over the plane either. Its variable type is `"singular"`.

## Slicing: fixing the total of two flows

Suppose two rivers feed a lake, and the question is how the inflow splits between them *given* that their total is extreme --- say, its 200-year value. That is the joint distribution of the two flows, sliced along the line where they add up to that value.

With `prob()`, that is a condition in `given`: the total taking a value.

```{r}
flows
```

The 200-year total is a return level of the total's distribution. For Normal flows the total is Normal too, with mean and variance from those of the flows:

```{r}
total <- dst_norm(
  mean = sum(mean(flows)),
  sd = sqrt(sum(variance(flows)))
)
s200 <- eval_return(total, at = 200)
s200
```

Now the chance that the first river carries more than half of that total, and the chance that it carries at most 60, 70 or 80 of it:

```{r}
prob(flows, r1 > r2, given = r1 + r2 == s200)
prob(flows, r1 <= c(60, 70, 80), given = r1 + r2 == s200)
```

This is exact for the multivariate Normal and t, whose sums of multiples of variables are again Normal or t. Behind the scenes, the total is added as a third variable and conditioned on. The three variables have a singular covariance --- the total is determined by the flows --- and live on a plane, so their variable type is `"singular"`:

```{r}
a <- rbind(r1 = c(1, 0), r2 = c(0, 1), s = c(1, 1))
trio <- dst_mv_norm(
  mean = setNames(as.numeric(a %*% mean(flows)), rownames(a)),
  cov = a %*% variance(flows) %*% t(a)
)
vtype(trio)
```

To keep the slice itself as a distribution, rather than asking it for probabilities, condition `trio` on `s` with `conditional()` from distplyr. The result still has two variables, `r1` and `r2`, but lives on a line, with no density over the plane; `marginal()` reduces it to one flow, which has one.

## Your own multivariate distribution

`distribution()` builds a multivariate distribution just as it builds a univariate one. Give `.support` a multivariate support, and write the representations as functions of one argument per variable, in order.

```{r}
storm <- distribution(
  density = function(x, y) dnorm(x) * dexp(y),
  cdf = function(x, y) pnorm(x) * pexp(y),
  .support = support_product(
    x = continuous(),
    y = continuous(c(0, Inf))
  )
)
```

Whatever is not stated is worked out where it can be: the survival function from the CDF, marginals by setting the other variables to infinity or integrating them out, and conditionals by dividing densities.

```{r}
eval_bi_survival(storm, x = 0, y = 1)
eval_bi_cdf(storm, x = 0.3, y = 1, known = "x")
mean(marginal(storm, "y"))
```

A distribution can also state its own marginals and conditionals, as the multivariate Normal does, when it has closed forms for them: see the "Several variables" section of `?distribution`.

## Not yet

A few things are not supported yet:

- Supports with a mix of kinds within one joint, such as a count paired with an amount, can be described, but such a distribution has no joint density or PMF to evaluate.
- Moments beyond the mean and covariance matrix.
- Plotting, and `enframe_*()` functions, for several variables.
- Events on combinations of variables other than sums of multiples of them (such as `x * y > 1`), except for distributions on finitely many points.
