---
title: "Multivariate Distributions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multivariate Distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
has_mvtnorm <- requireNamespace("mvtnorm", quietly = TRUE)
```

```{r setup}
library(distionary)
```

A distribution can describe several variables at once: the flows of two rivers on the same day, say, or a storm's duration together with its depth. This vignette shows how to build such distributions, how to evaluate them, and how to condition them --- including slicing one along a line, such as fixing the total of two flows.

Multivariate distributions are new in `distionary`, and still experimental. The names and behaviour described here may change.

## Building one

The multivariate Normal comes in two forms: `dst_mv_norm()` for any number of variables, specified by a mean vector and covariance matrix, and `dst_bi_norm()`, a shortcut for two, specified by standard deviations and a correlation.

```{r}
d <- dst_bi_norm(mean = c(0, 1), sd = c(1, 2), cor = 0.6)
d
```

Its variables have names, and `dimension()` counts them.

```{r}
variables(d)
dimension(d)
```

Name them yourself through the names of the mean:

```{r}
flows <- dst_mv_norm(
  mean = c(r1 = 50, r2 = 80),
  cov = matrix(c(100, 60, 60, 225), 2)
)
variables(flows)
```

The other built-in family is the empirical distribution, which places probability on each observed point --- each row of a data set, taken as a whole.

```{r}
obs <- data.frame(
  flow = c(10, 12, 15, 11, 12),
  depth = c(1.0, 1.5, 2.0, 1.1, 1.5)
)
e <- dst_bi_empirical(flow, depth, data = obs)
e
```

`dst_mv_empirical()` does the same for any number of variables, taking a list of vectors --- a data frame being one.

There is also the multivariate Student t, `dst_mv_t()` (and `dst_bi_t()`), whose heavier tails make its variables extreme together more often than the Normal's. Everything below that works for the Normal --- including slicing --- works for it too.

## Evaluating one

A univariate distribution is evaluated at a vector of values. A multivariate one needs a value for each variable, so there are two families of evaluators, much as `purrr` has `map2()` and `pmap()`:

- `eval_bi_*(d, x, y)` for two variables, one vector each;
- `eval_mv_*(d, l)` for any number, a list of vectors.

```{r, eval = has_mvtnorm}
eval_bi_cdf(d, x = 0, y = 1)
eval_mv_cdf(d, list(0, 1))
eval_bi_density(d, x = 0:2, y = 1)
```

The vectors are recycled to a common length, and element `i` of the result is the representation at the `i`th point. A named list is matched to the variables by name, so its order does not matter; a data frame works just as well. That includes the draws from `realise()`, which come back as a data frame:

```{r}
set.seed(1)
draws <- realise(flows, n = 3)
draws
eval_mv_density(flows, draws)
```

The functions for one variable refuse a multivariate distribution, and say what to use instead:

```{r, error = TRUE}
eval_cdf(d, 0)
```

### Survival, and other corners

The survival function is the probability that *every* variable exceeds its value, \(P(X > x, Y > y)\). That is not one minus the CDF, which is the probability that *at least one* does; the two agree only for a single variable.

```{r, eval = has_mvtnorm}
eval_bi_survival(d, x = 0, y = 1)
1 - eval_bi_cdf(d, x = 0, y = 1)
```

For any other combination of inequalities --- one variable below its value and the other above, say --- use `prob_bi()` or `prob_mv()`:

```{r, eval = has_mvtnorm}
prob_bi(d, x = 0, y = 1, ineq = c("<=", ">"))
```

Strict and non-strict inequalities differ only where a variable has atoms:

```{r}
prob_bi(e, x = 12, y = 1.5, ineq = c("<", "<="))
prob_bi(e, x = 12, y = 1.5, ineq = c("<=", "<="))
```

## Marginals and conditionals

`marginal()` gives the distribution of some of the variables, ignoring the rest. With one variable, it is an ordinary univariate distribution, to be evaluated with the usual functions.

```{r}
r2 <- marginal(flows, "r2")
r2
eval_quantile(r2, at = 0.99)
```

To evaluate a representation *conditional* on some of the variables, say which ones are known with `given`. The arguments stay in the same order either way: every variable still gets a value, and `given` says which of those values are known rather than evaluated at. The variables in `given` are the ones on the right of the bar.

```{r, eval = has_mvtnorm}
# P(Y <= y | X = 2), for y = 0, 1, 2
eval_bi_cdf(d, x = 2, y = 0:2, given = "x")
```

To have the conditional distribution itself, as a distribution to evaluate, draw from or condition further, use `conditional()` from the [distplyr](https://distplyr.probaverse.com) package, which makes new distributions out of existing ones.

## Supports of several variables

Every distribution says where it places its probability, through its support (see `vignette("support")`). The simplest multivariate support is a *product*: every value of one variable paired with every value of each of the others, as `expand.grid()` pairs them.

```{r}
support_product(
  storms = discrete(natural0()),
  depth = continuous(c(0, Inf))
)
```

An empirical distribution's support is not a product, though: it is the set of points observed, not every pairing of an observed flow with an observed depth. `discrete()` takes such a set as a data frame.

```{r}
support(e)
```

For anything else, describe the support as the image of a simpler one under a map. The triangle \(0 < y < x\) is where \((x, z) \mapsto (x, xz)\) sends the strip \(x > 0\), \(0 < z < 1\):

```{r}
strip <- support_product(
  x = continuous(c(0, Inf)),
  z = continuous(c(0, 1))
)
support_map(
  strip,
  fun = function(x, z) list(x, x * z),
  margins = list(
    x = continuous(c(0, Inf)),
    y = continuous(c(0, Inf))
  )
)
```

A general map has to be told what values each variable reaches. A linear one does not, and `support_affine()` works them out. Here is the line \(x + y = 100\), as the image of the real line under \(t \mapsto (50 + t, 50 - t)\):

```{r}
line <- support_affine(
  continuous(),
  shift = c(x = 50, y = 50),
  matrix = rbind(1, -1)
)
line
```

A line has two variables but spans only one dimension. A distribution on it puts no probability on any single point, but has no density over the plane either. Its variable type is `"singular"`.

## Slicing: fixing the total of two flows

Suppose two rivers feed a lake, and the question is how the inflow splits between them *given* that their total is extreme --- say, its 200-year value. That is the joint distribution of the two flows, sliced along the line where they add up to that value.

Slicing is conditioning, on a variable that is a function of the others. So add the total as a third variable, and condition on it. For Normal flows this is exact, because a linear combination of Normal variables is again Normal: if the flows have mean \(\mu\) and covariance \(\Sigma\), then \(A\) times them has mean \(A\mu\) and covariance \(A \Sigma A^\top\).

```{r}
mu <- c(r1 = 50, r2 = 80)
sigma <- matrix(c(100, 60, 60, 225), 2)
a <- rbind(r1 = c(1, 0), r2 = c(0, 1), s = c(1, 1))
trio <- dst_mv_norm(
  mean = setNames(as.numeric(a %*% mu), rownames(a)),
  cov = a %*% sigma %*% t(a)
)
trio
```

The covariance of the three is singular --- the total is determined by the flows --- so the three live on a plane, and their variable type is `"singular"`.

The 200-year total is a return level of the total's marginal distribution:

```{r}
s200 <- eval_return(marginal(trio, "s"), at = 200)
s200
```

Conditioning on it slices the distribution: the two flows now live on the line \(r_1 + r_2 = s_{200}\), so fixing one fixes the other. Evaluate the three-variable distribution with `given = "s"`, leaving the other flow unrestricted (at `Inf` for a CDF, `-Inf` for a survival function). Here is the chance that the first river carries at most 60, 70 or 80 of the total:

```{r, eval = has_mvtnorm}
eval_mv_cdf(
  trio,
  list(r1 = c(60, 70, 80), r2 = Inf, s = s200),
  given = "s"
)
```

And the chance that it carries more than half:

```{r, eval = has_mvtnorm}
eval_mv_survival(
  trio,
  list(r1 = s200 / 2, r2 = -Inf, s = s200),
  given = "s"
)
```

distplyr's `conditional()` gives the slice as a distribution of its own. It still has two variables, but its variable type is `"singular"`: it has no density over the plane. Reduce it to one flow with `marginal()` to get one with a density, as in `marginal(conditional(trio, given = c(s = s200)), "r1")`.

## Your own multivariate distribution

`distribution()` builds a multivariate distribution just as it builds a univariate one. Give `.support` a multivariate support, and write the representations as functions of one argument per variable, in order.

```{r}
storm <- distribution(
  density = function(x, y) dnorm(x) * dexp(y),
  cdf = function(x, y) pnorm(x) * pexp(y),
  .support = support_product(
    x = continuous(),
    y = continuous(c(0, Inf))
  )
)
```

Whatever is not stated is worked out where it can be: the survival function from the CDF, marginals by setting the other variables to infinity or integrating them out, and conditionals by dividing densities.

```{r}
eval_bi_survival(storm, x = 0, y = 1)
eval_bi_cdf(storm, x = 0.3, y = 1, given = "x")
mean(marginal(storm, "y"))
```

A distribution can also state its own marginals and conditionals, as the multivariate Normal does, when it has closed forms for them: see the "Several variables" section of `?distribution`.

## Not yet

A few things are not supported yet:

- Supports with a mix of kinds within one joint, such as a count paired with an amount, can be described, but such a distribution has no joint density or PMF to evaluate.
- Moments beyond the mean and covariance matrix.
- Plotting, and `enframe_*()` functions, for several variables.
- Conditioning on an event such as \(X > x\), rather than on a value. Divide two probabilities from `prob_mv()` instead: \(P(Y > y \mid X > x) = P(X > x, Y > y) / P(X > x)\).
