---
title: "The Support of a Distribution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Support of a Distribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(distionary)
```

This vignette explains what a distribution's _support_ is in `distionary`, how to build one, and why it's worth telling the package about it.

## What a Support Is

A distribution's **support** is the set on which it places probability. Every distribution splits that set into two parts:

- an **atomic** part -- individual points, each carrying positive probability mass; and
- a **continuous** part -- one or more intervals, over which probability is spread as a density.

This split is the Lebesgue decomposition, and it's the reason a distribution is called discrete (atoms only), continuous (intervals only), or mixed (both). In `distionary`, a support is an object holding these two parts, and the variable type is _derived_ from it rather than declared.

The reason to carry this structure around is practical: numeric routines need to know where the atoms are. A quantile function has to land exactly on an atom rather than near it, and a mean has to sum over atoms rather than integrate across them. Recording the support means these calculations can be done correctly without having to guess.

## Building a Support

Three constructors build supports, one for each variable type.

### Continuous supports

`continuous()` takes intervals, each a length-2 numeric vector `c(lower, upper)`.

```{r}
continuous(c(0, Inf))
```

With no arguments at all, it defaults to the whole real line.

```{r}
continuous()
```

Several intervals give a union, which is sorted and merged into a canonical form. Notice that the overlapping pieces below collapse into one:

```{r}
continuous(c(3, 4), c(0, 1), c(0.5, 2))
```

Intervals are written as closed, but their endpoints carry no probability either way, a single point having measure zero. An interval is really neither open nor closed. A point that genuinely carries mass is an atom, and belongs in the atomic part instead. Each interval must have `lower < upper`.

### Discrete supports

`discrete()` takes the atoms. Finitely many atoms can be given as a plain numeric vector, which is sorted for you.

```{r}
discrete(c(3.5, 1.2, 6.7))
```

Distributions like the Poisson have infinitely many atoms, which can't be written out as a vector. These come from the `discretes` package, whose series constructors `distionary` re-exports for convenience.

```{r}
discrete(natural0())
```

```{r}
discrete(integers())
```

`arithmetic()` describes evenly spaced atoms, given a representative point and a spacing. It runs forever in both directions unless you say otherwise:

```{r}
discrete(arithmetic(0, 0.5))
```

```{r}
discrete(arithmetic(0, 0.5, n_left = 0, n_right = 4))
```

### Mixed supports

`mixed()` takes both parts, named. Here is an atom at zero sitting at the edge of a continuous tail:

```{r}
mixed(atoms = 0, continuous = c(0, Inf))
```

An atom is allowed to coincide with an interval endpoint, as it does here -- the atom is simply tracked separately, and its mass is not double-counted against the density. Because the variable type is derived from the parts, `mixed()` insists that _both_ of them be non-empty; use `discrete()` or `continuous()` for the pure cases.

### The empty support

There is a fourth constructor, for a support with nothing in it at all.

```{r}
empty_support()
```

No distribution has an empty support -- probability has to go somewhere -- and `distribution()` refuses one.

```{r, error = TRUE}
distribution(cdf = stats::pnorm, .support = empty_support())
```

It exists so that operations on supports are _closed_. Restricting a support to a region it doesn't reach has to return something, and the empty support is that something. Without it, every operation that can come up empty would need a special case, and the thing it returned in that case would have to mean something else.

Its variable type, shown when it prints, is `"empty"`, which says there is nowhere to place probability at all. That is a different thing from a distribution simply not saying where its probability lives, which `distribution()` no longer permits.

```{r}
is_empty_support(empty_support())
```

## Retrieving a Support

The built-in families all carry a support, which `support()` retrieves.

```{r}
support(dst_pois(3))
support(dst_gamma(shape = 2, rate = 1))
```

The two parts can be pulled out individually. `atoms()` returns a `discretes` object, and `continuous_part()` returns a two-column matrix of intervals.

```{r}
atoms(mixed(atoms = 0, continuous = c(0, Inf)))
continuous_part(continuous(c(0, 1), c(3, 4)))
```

Both accept a distribution as well as a support, so there's no need to unwrap first.

```{r}
atoms(dst_binom(5, 0.3))
```

The variable type and the range are both derived from the support, so they always agree with it.

```{r}
vtype(dst_pois(3))
range(dst_pois(3))
```

A support has a range of its own, being the two outermost points it reaches.

```{r}
range(continuous(c(0, 1), c(3, 4)))
```

Gaps aren't represented there. The support above places no probability between 1 and 3, but its range spans that stretch all the same.

Finally, `is_support()` tests whether an object is a support.

```{r}
is_support(continuous(c(0, 1)))
```

## Manipulating a Support

Supports can be combined, cut down, and mapped through functions. Every one of these operations takes supports and gives back a support, so they compose freely.

Two supports can be unioned. Overlapping or touching intervals are merged, and the atoms pooled.

```{r}
support_union(continuous(c(0, 1)), continuous(c(0.5, 3)))
```

An atom lying inside another support's continuous part stays an atom. The two carry different kinds of probability, and neither absorbs the other.

```{r}
support_union(discrete(3), continuous(c(0, 10)))
```

`support_restrict()` cuts a support down to an interval. The `include_from` and `include_to` arguments apply to the atoms only, since an interval endpoint carries no probability either way.

```{r}
support_restrict(discrete(natural0()), to = 4)
```

```{r}
support_restrict(discrete(natural0()), to = 4, include_to = FALSE)
```

Restricting to somewhere the support doesn't reach is where the empty support earns its keep.

```{r}
support_restrict(continuous(c(0, 1)), from = 5, to = 6)
```

A support can also be pushed through a strictly monotonic function. Shifting and scaling are common enough to have their own forms.

```{r}
support_shift(continuous(c(0, 1)), by = 5)
support_scale(discrete(natural0()), by = 2)
```

A decreasing map reverses each interval, which is taken care of for you.

```{r}
support_scale(continuous(c(1, 2)), by = -1)
```

Reciprocals are worth a closer look, because `1 / x` isn't monotonic across zero. Each side of zero is mapped separately and the two results unioned, which is why the support below comes back in two pieces.

```{r}
support_reciprocal(continuous(c(-2, 4)))
```

For anything else, `support_transform()` takes the map together with its inverse.

```{r}
support_transform(
  continuous(c(0, Inf)),
  fun = exp, inv = log,
  domain = c(0, Inf), range = c(1, Inf)
)
```

Atoms can be added and removed one at a time. This leaves the continuous part alone, so removing an atom that sits on an interval leaves the interval whole.

```{r}
support_add_atoms(continuous(c(0, Inf)), 0)
```

Finally, two functions ask whether a value belongs to a support. `support_contains()` counts both parts, whereas `support_has_atom()` counts only the atoms --- making it the one to reach for when what matters is whether a point carries positive probability.

```{r}
s <- mixed(atoms = 0, continuous = c(2, 5))
support_contains(s, at = c(0, 1, 3, 9))
support_has_atom(s, at = c(0, 1, 3, 9))
```

## Giving a Distribution a Support

Supports are attached to user-defined distributions through the `.support` argument of `distribution()`, described in the [Specifying Your Own Distribution](./specify-user-defined.html) vignette.

```{r}
my_normal <- distribution(
  cdf = stats::pnorm,
  density = stats::dnorm,
  .support = continuous(c(-Inf, Inf))
)
vtype(my_normal)
```

One rule is worth knowing about. A bare numeric vector is not accepted as a support, because it is ambiguous: `c(0, Inf)` could just as easily mean an interval from zero to infinity as it could mean two atoms. You have to name which one you mean.

```{r, error = TRUE}
distribution(cdf = stats::pnorm, .support = c(0, Inf))
```

Inside `discrete()` and `mixed()` the ambiguity is already resolved by the argument name, which is why a bare vector is fine there.

A support is not optional. It is the one thing `distionary` cannot work out from the representations you give it: a CDF says how much probability lies below a point, but not where the atoms are, nor where the distribution ends.

```{r, error = TRUE}
distribution(cdf = stats::pnorm, density = stats::dnorm)
```

The older `.vtype` argument, which took a string such as `"continuous"`, is superseded and now has no effect: the variable type is always derived from the support. Naming a type was never the same as saying where the probability is, which is why it cannot stand in for one.

## Why It Matters

The payoff shows up on distributions that have atoms. Consider a model for daily rainfall: no rain at all with probability 0.3, and an exponential amount otherwise. This is a genuinely mixed distribution, with an atom at zero and a density on the positive half-line.

```{r}
p0 <- 0.3
rate <- 1 / 5
rainfall <- distribution(
  cdf = function(x) {
    ifelse(x < 0, 0, p0 + (1 - p0) * stats::pexp(x, rate))
  },
  density = function(x) {
    ifelse(x <= 0, 0, (1 - p0) * stats::dexp(x, rate))
  },
  pmf = function(x) {
    ifelse(x == 0, p0, 0)
  },
  .support = mixed(atoms = 0, continuous = c(0, Inf)),
  .name = "Rainfall"
)
vtype(rainfall)
```

Only the CDF, density, and PMF were specified. The quantile function is derived, and because the support says there is an atom at zero, it lands on zero exactly for every probability at or below 0.3 instead of returning something merely close.

```{r}
eval_quantile(rainfall, at = c(0.1, 0.3, 0.5, 0.9))
```

The mean is derived too, by summing over the atoms and integrating over the intervals separately. Here the answer is `(1 - p0) / rate`, which is 3.5.

```{r}
mean(rainfall)
variance(rainfall)
```

The same applies to a purely discrete distribution defined only by its CDF and PMF. Quantiles come back as exact atoms:

```{r}
my_poisson <- distribution(
  cdf = function(x) stats::ppois(x, 2.5),
  pmf = function(x) stats::dpois(x, 2.5),
  .support = discrete(natural0()),
  .name = "My Poisson"
)
eval_quantile(my_poisson, at = c(0.05, 0.3, 0.5, 0.9, 0.99))
```

Note that the support here has infinitely many atoms, and the mean is still computed, by walking outward through the atoms until the remaining contributions are negligible.

```{r}
mean(my_poisson)
```

## Current Scope

Two things are worth knowing about where supports stand today.

A support describes a distribution on the real line. Multivariate supports would be composed from univariate ones, and the object already reserves a place to record its dimension, but that composition isn't built yet.

There is no intersection. It hasn't been needed --- `support_restrict()` already covers cutting a support down to an interval, which is the case that comes up in practice --- so it waits until something calls for it.
